# Only healthy samples

```{r}
library("pheatmap")
library("dplyr")
library("viridis")
library("ggpubr")

# vectors to filter out healthy samples from Gupta
samples <- read.table('data/otutable_metadata.csv',
                   sep = ",", header = TRUE, row.names = 1) %>%
  filter(study_name == "GuptaA_2019") %>% rownames()
gupta_all <- gsub("GupDM_", "", samples)
gupta_healthy <-  gupta_all[grep("^H", gupta_all)]


# load meta for species for higher taxonomical annotation
metasp <- read.table('data/metaspecies.csv',
                     sep = ",", header = TRUE, row.names = 1)
annot_phyla <- metasp['phylum']

# annotation for samples
annot_condition <- data.frame(row.names = gupta_all,
                              'стан' = ifelse(substr(gupta_all, 1, 1) == "H", "здорові", "рак"))

# to add titles in a function
titles <- list("prepped" = "Сирі дані", "css" = "CSS", "blom" = "Blom",
                "clr" = "CLR", "combat" = "ComBat")
```


```{r}
plotheat_healthy_samples <- function(method, addlegend = FALSE){
  pathtoftable <- paste0('camp_normalization_out/normalization/otutable_', method, '.csv')
  ftable <- read.table(pathtoftable,
                       sep = ",", header = TRUE, row.names = 1)
  colnames(ftable) <- gsub("GupDM_", "", colnames(ftable))
  
  # sort out the healthy samples
  ftable <- ftable[, gupta_healthy]

  corr_matrix <- cor(ftable)
  pheatmap(corr_matrix, show_colnames = FALSE, show_rownames = TRUE,
           color = magma(256, begin = .15, direction = -1), border_color = NA,
           legend = addlegend, main = titles[method])
}

```

```{r}
plotheat_all_samples <- function(method, addlegend = FALSE){
  pathtoftable <- paste0('camp_normalization_out/normalization/otutable_', method, '.csv')
  ftable <- read.table(pathtoftable,
                       sep = ",", header = TRUE, row.names = 1)
  colnames(ftable) <- gsub("GupDM_", "", colnames(ftable))
  
  ftable <- ftable[, gupta_all]
  
  corr_matrix <- cor(ftable)
  pheatmap(corr_matrix, show_colnames = FALSE, show_rownames = FALSE,
           color = magma(256, begin = .15, direction = -1), border_color = NA,
           legend = addlegend, main = titles[method],
           annotation_col = annot_condition)
}
```

```{r}
plotheat_species <- function(method, healthy = TRUE){
  pathtoftable <- paste0('camp_normalization_out/normalization/otutable_', method, '.csv')
  ftable <- read.table(pathtoftable,
                       sep = ",", header = TRUE, row.names = 1)
  colnames(ftable) <- gsub("GupDM_", "", colnames(ftable))
  
  if (healthy){ # sort out the healthy samples
    ftable <- ftable[, gupta_healthy]
  } else{ # sort out all Gupta samples
    ftable <- ftable[, gupta_all]
  }
  
  # transpose the feature table
  ftable <- data.frame(t(ftable), check.names = FALSE)
  # filter out the species with `the same` counts among all samples
  ftable <- ftable[sapply(ftable, function(x) length(unique(x))>1)]
  
  # subset the metaspecies to match the ftable
  #metasp <- metasp[rownames(metasp) %in% colnames(ftable), ]
  
  
  corr_matrix <- cor(ftable)
  pheatmap(corr_matrix, show_colnames = FALSE, show_rownames = FALSE,
           color = magma(256, begin = .15, direction = -1), border_color = NA,
           legend = TRUE, main = titles[method],
           annotation_col = annot_phyla)
}
```

## Correlation in species
```{r}
prep <- plotheat_species("prepped")
css <- plotheat_species("css")
clr <- plotheat_species("clr")
combat <- plotheat_species("combat")
blom <- plotheat_species("blom")

ggarrange(prep[[4]],
          ggarrange(css[[4]], clr[[4]], blom[[4]], combat[[4]]), 
          widths = c(1, 2), labels = "AUTO") %>% 
  ggexport(filename = "figures/heatmap_healthy-species.png",
           width = 1960, height = 980, res = 150)
```

## Correlation in samples

```{r}
prep <- plotheat_samples("prepped", addlegend = TRUE, healthy = FALSE)
css <- plotheat_samples("css", healthy = FALSE)
clr <- plotheat_samples("clr", healthy = FALSE)
combat <- plotheat_samples("combat", healthy = FALSE)
blom <- plotheat_samples("blom", healthy = FALSE)

ggarrange(prep[[4]],
          ggarrange(css[[4]], clr[[4]], blom[[4]], combat[[4]]), 
          widths = c(1, 2), labels = "AUTO") %>% 
  ggexport(filename = "figures/heatmap_healthy-samples.png",
           width = 1960, height = 980, res = 150)

```

# All samples

```{r}
prep <- plotheat_all_samples("prepped", addlegend = TRUE)
css <- plotheat_all_samples("css")
clr <- plotheat_all_samples("clr")
combat <- plotheat_all_samples("combat")
blom <- plotheat_all_samples("blom")

ggarrange(prep[[4]],
          ggarrange(css[[4]], clr[[4]], blom[[4]], combat[[4]]), 
          widths = c(1, 2), labels = "AUTO") %>% 
  ggexport(filename = "figures/heatmap_all-samples.png",
           width = 1960, height = 1100, res = 150)

```

# Samples separately

```{r}
# List of dataframes
prep <- read.table(paste0('camp_normalization_out/normalization/otutable_', 'prepped', '.csv'),
                       sep = ",", header = TRUE, row.names = 1)
css <- read.table(paste0('camp_normalization_out/normalization/otutable_', 'css', '.csv'),
                       sep = ",", header = TRUE, row.names = 1)
clr <- read.table(paste0('camp_normalization_out/normalization/otutable_', 'clr', '.csv'),
                       sep = ",", header = TRUE, row.names = 1)
blom <- read.table(paste0('camp_normalization_out/normalization/otutable_', 'blom', '.csv'),
                       sep = ",", header = TRUE, row.names = 1)
combat <- read.table(paste0('camp_normalization_out/normalization/otutable_', 'combat', '.csv'),
                       sep = ",", header = TRUE, row.names = 1)

df_list <- list(prep, css, clr, blom, combat)

```

```{r}
methods <- c("prep", "css", "clr", "blom", "combat")

plotheat_methods <- function(sample){
 # extract one sample columns from each dataframe
  cols_combined <- lapply(seq_along(df_list), function(i) {
    col <- df_list[[i]][sample]
    return(col)
  })
  
  # combine  columns into a single dataframe
  cols_combined <- data.frame(do.call(cbind, cols_combined))
  colnames(cols_combined) <- c("Сирі дані", "СSS", "CLR", "Blom", "ComBat")
  corr_matrix <- cor(cols_combined)
  p <- pheatmap(corr_matrix, show_colnames = TRUE, show_rownames = TRUE,
             color = magma(256, begin = .15, direction = -1), border_color = NA,
             legend = TRUE, cluster_cols = FALSE, cluster_rows = FALSE, main = sample) 
  return(p[[4]])
}
```

```{r}
plots <- lapply(samples[1:10], plotheat_methods)

# Arrange the plots using ggarrange
ggarrange(plotlist = plots, ncol = 5, nrow = 2) %>% 
  ggexport(filename = "figures/heatmap_methods.png",
           width = 1960, height = 768, res = 150)
```
The graph shows the correlation between 4 normalization methods and raw data for several healthy samples in Gupta 2029. The general trend is strong correlation between CSS and Combat, with CSS gravitating towards the row data The least correlation is observed between Blom and raw data and CLR with both ComBat and Blom. Overall, the tendencies described might be visually observed on the heatmaps that illustrate correlation between samples.
